**Best practices for applying machine learning to bacterial 16S rRNA gene sequencing data**

### 2019-08-26 submission/manuscript.Rmd
add changes after 4th meeting

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism, but are the product of multiple bacterial populations. Whereas traditional statistical approaches are useful for identifying those cases where a single organism can drive a process, ML methods offer the ability to use many features of these communities to identify those cases. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those populations that are responsible for the classification. Because application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models [Table 1].

@@

Even with interpretable models such as L2-regularized logistic regression, there are potential pitfalls when it comes to identifying biomarkers of a disease. Human-associated microbial communities have complex correlation structures. This can hinder our ability to reliably interpret models [@dormann_collinearity:_2013]. In this study we used two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. The feature weights of correlated OTUs are influenced by and dependent on one another. To avoid misinterpreting the models, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. To avoid misinterpreting models due to correlation structures of the communities when we used permutation importance, we grouped correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features to identify the true underlying factors when making a classification.

In this study, we established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome. We used a held-out test set and we performed the initial 80%-20% random datasplit to create the held-out test set 100 times in our ML pipeline. The models we built with this pipeline were generalizable despite the high number of features microbiome datasets usually have.

>>>>

There is a growing awareness that many human diseases and environmental processes are not driven by a single organism, but are the product of multiple bacterial populations. Whereas traditional statistical approaches are useful for identifying those cases where a single organism can drive a process, ML methods offer the ability to use many features of these communities to identify those cases. If it is possible to classify communities reliably, then ML methods also offer the ability to identify those populations that are responsible for the classification. Because the application of ML in microbiome studies is still in its infancy, we developed a framework to train rigorous, transparent and reproducible ML models where we pointed out potential pitfalls and ways to avoid them [Table 2].

@@

Even with interpretable models such as L2-regularized logistic regression, there are potential pitfalls when it comes to identifying biomarkers of a disease. Human-associated microbial communities have complex correlation structures. This can hinder our ability to reliably interpret models [@dormann_collinearity:_2013]. In this study we used two different interpretation methods; ranking each OTU by (1) their absolute weights in the trained models and (2) their impact on the predictive performance based on permutation importance. The feature weights of correlated OTUs are influenced by one another. For example if one OTU is highly correlated with another OTU, only one of these OTUs might have a large feature weight in the regression model. To avoid misinterpreting the models and missing important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs as well. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we avoided misinterpreting models due to correlation structures of the communities by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other however, we can reduce the correlation threshold to further investigate the relationships among features.

We established a pipeline that can be generalized to any modeling method that uses 16S rRNA sequence counts to predict a binary health outcome. We performed an initial 80%-20% random datasplit to create an held-out test set which we used to evaluate model performance. We repeated this datasplit 100 times to get a robust reading of predictive performance. We tuned our model hyperparameters in a five-fold cross-validation. The models we built with this pipeline were generalizable despite the high number of features microbiome datasets usually have.
